Approximation as Differentiation: Notes

### 2025-05-16 main/formal-structures.tex
Going to introduce enrichment of LatGal earlier.

@@

\noindent In $\Fam(\cat{C})$ the objects $(X, \partial X)$ are thus sets $X$ paired with indexed families
$\partial X: X \to \cat{C}$ and morphisms $(f, \partial f): (X, \partial X) \to (Y, \partial Y)$ are functions
$f: X \to Y$ paired with morphisms $\partial f: \partial X \to \partial Y \comp f$ in $\Func{X}{\cat{C}}$.

@@

Now consider the total category $\Fam(\LatGal)$, which is our basic setting for interpreting first-order
programs for Galois slicing. This has as objects $(X, \partial X)$, all pairs of a set $X$ and and for every
$x \in X$, a bounded lattice $\partial X(x)$, and as morphisms $(X, \partial X) \to (Y, \partial Y)$, all
pairs $(f, \partial f)$ of a function $f: X \to Y$ and for every $x \in X$, a Galois connection $\partial
f(x): \partial X(x) \to \partial Y(f(x))$, as required. The pullback (reindexing) along $f$ in the composition
$(g, \partial g) \comp (f, \partial f) = (g \comp f, \partial g \comp f \comp \partial f)$ selects the
appropriate lattice of approximations and Galois connection at each point, using the baseline (unapproximated)
function $f$.

@@

An important observation about the semilattice homomorphisms that appear in adjoint pairs in $\LatGal$ is that
the pointwise lifting of $\meet$ to meet-semilattice homomorphisms $f, g: X \to Y$, where $(f \meet g)(x) =
f(x) \meet g(x): X \to Y$, gives a notion of ``addition'' for those maps which is associative and commutative,
and which has a ``zero'' homomorphism $0_{X,Y}: X \to Y$ given by the constant map sending any $x \in X$ to
$0_Y$, acting as a unit for addition of homomorphisms. The same can be said for the join-preserving maps from
$Y$ to $X$, which also have a commutative monoid structure; Galois connections then inherit this structure
component-wise.

We now say more generally when this so-called \emph{semi-additive} structure on morphisms arises and explain
its role in generalising CHAD from automatic differentiation to providing a semantics for Galois slicing as
well.

>>>>

\todo{Introduce that $\LatGal$ is $\SemiLat$-enriched -- the hom-set $\LatGal(X)(Y)$ is has an (idempotent) addition of morphisms and a zero morphism. Moreover it has biproducts -- product of lattices $X \times Y$ is also a coproduct, although formal definition will come later.}

% the pointwise lifting of $\meet$ to meet-semilattice homomorphisms $f, g: X \to Y$, where $(f \meet g)(x) =
% f(x) \meet g(x): X \to Y$, gives a notion of ``addition'' for those maps which is associative and commutative,
% and which has a ``zero'' homomorphism $0_{X,Y}: X \to Y$ given by the constant map sending any $x \in X$ to
% $0_Y$, acting as a unit for addition of homomorphisms. The same can be said for the join-preserving maps from
% $Y$ to $X$, which also have a commutative monoid structure; Galois connections then inherit this structure
% component-wise.

@@

\noindent In $\Fam(\cat{C})$ the objects $(X, \partial X)$ are thus sets $X$ paired with indexed families
$\partial X: X \to \cat{C}$ and morphisms $(f, \partial f): (X, \partial X) \to (Y, \partial Y)$ are functions
$f: X \to Y$ paired with morphisms $\partial f: \partial X \to \partial \reindex{Y}{f}$ in
$\Func{X}{\cat{C}}$.

@@

Now consider the total category $\Fam(\LatGal)$, which is our basic setting for interpreting first-order
programs for Galois slicing. This has as objects $(X, \partial X)$, all pairs of a set $X$ and and for every
$x \in X$, a bounded lattice $\partial X(x)$, and as morphisms $(X, \partial X) \to (Y, \partial Y)$, all
pairs $(f, \partial f)$ of a function $f: X \to Y$ and for every $x \in X$, a Galois connection $\partial
f(x): \partial X(x) \to \partial Y(f(x))$, as required. The pullback (reindexing) along $f$ in the composition
$(g, \partial g) \comp (f, \partial f) = (g \comp f, \reindex{\partial g}{f} \comp \partial f)$ selects the
appropriate lattice of approximations and Galois connection at each point, using the baseline (unapproximated)
function $f$.

@@

We noted that $\LatGal$ is $\CMon$-enriched (actually $\SemiLat$-enriched) and moreover that the product of
lattices is also a coproduct, making $\LatGal$ into a category with \emph{biproducts}. In fact having
biproducts is sufficient to ensure $\CMon$-enrichment; such a category is called \emph{semi-additive}.
